---
title: "Pedogenesis_Data_Project"
author: "Phoebe Teskey"
date: "2026-04-30"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

***Soil Data Project*** The attached spreadsheets contain 1) summary of
the PDI and WM PDI values for the McAuliffe 1994 paper and the SRER soil
chronosequence, and 2) compiled long-term vegetation transect from the
SRER that includes data for mean annual precipitation (MAP) and
surficial geologic map unit (MapUnit) that corresponds with estimated
landscape age. This assignment will consist of several components.

***Component 1 - McAuliffe Paper***

1)  How do the calculated PDI and WM PDI values relate to vegetation
    abundance, diversity, structural form across the three fan complexes
    detailed in the paper? Were your hypotheses from the McAuliffe
    reading assignment confirmed?

PDI values can be used as a proxy for determining and understanding
vegetation composition. A higher PDI value indicates older, more
developed soils, which tend to have higher clay content and less water
penetration, selecting for shallow rooted plants such as grasses.
Intermediate PDI values are moderately developed, with some clay
accumulation and some deeper water infiltration possible. We see a high
plant diversity here, as this soil can support both shallow rooted and
some deep-rooted plants. Woody shrubs, perennial grass and mesquite are
common. The lower PDI values indicate the least developed soils, with
low clay content and deep, rapid water infiltration. This selects for
deeply rooted plants such as mesquite and some shrubs. However,
McAuliffe finds that “Vegetation patterns are not continuous along a
gradient and instead show strong spatial heterogeneity tied to
geomorphic surfaces”. He found that soil age does not necessarily inform
PDI values. The paper exemplifies this. The Tucson Mountains sites
exhibit both high and low PDI value sites, with low PDI sites having
more woody species and high PDI sites having more grass and shrubs
together. This loosely follows the general metric I described in the
first paragraph.

The Silverbell Mountains sites follow the vegetation/PDI metric most
closely, with woody species on young soils (low PDI) and mixed
grass/shrubs on middle-aged surfaces.

The Tortolitas are the least consistent. Some of the sites are very old,
but can support trees and shrubs, likely due to the high erosion at the
site, lending younger soil characteristics.

My hypotheses are here: 1. As clay and carbonates increase, so does
shrub dominance. 2. Plant diversity will peak at mid-aged soils with
both grasses and shrubs. 3. Higher clay accumulations will select for
shrubs because thicker, more sturdy roots are required to penetrate
clays and argillic horizons.

The first hypothesis is blatantly incorrect. Clay accumulation is
generally associated more with perennial grasses, not shrubs.

The second hypothesis follows the general PDI/vegetation metric seen in
the Tucson Mtns and Silverbells, but the Tortolitas provide an exception
to these general rules. As McAuliffe found, soil age is not always a
strong predictor of PDI values.

The third hypothesis is also false and is an iteration of the first. I
don’t know what I was thinking.

2)  How does erosion impact PDI and WM PDI on the different surfaces and
    what key soil features differ between the two surfaces that causes
    the change in profile development?

Erosion lends a lower PDI value to sites regardless of soil age. A site
with higher erosion has faster and deeper water infiltration, selecting
for plants that can access deeper water, like mesquite. This makes sense
in the Tortolitas, particularly because many of the surfaces there are
granitic, decomposing bedrock. The particle size is going to be larger,
clay content is lower, and in many ways, soil formation has been reset.

***Component 2 - SRER Chronosequence***

1)  Plot both PDI and WM PDI versus estimated soil age - using both the
    actual values and as a semi-log plot with soil age on a log scale.
    Provide detail on how both change relative to soil age as well as
    proposed explanations for the observed patterns.

```{r}
library(tidyverse)
SRER <- read.csv("../data/SRER2.csv")

SRER1 <- mutate(SRER, age_log = log(Age))

ggplot(SRER1, aes(x = age_log))+
  geom_line(aes(y = PDI), linewidth = 1) +
  theme_light() +
  labs(title = "PDI Values by Soil Age at the SRER", x = "Soil Age (log scale)", y = "PDI Value")
```

```{r}
ggplot(SRER1, aes(x = age_log))+
  geom_line(aes(y = WM_PDI), linewidth = 1) +
  theme_light() +
  labs(title = "WM_PDI Values by Soil Age at the SRER", x = "Soil Age (log scale)", y = "WM_PDI Value")
```

```{r}
McAuliffe <- read.csv("../data/McAuliffe.csv")

McAuliffe1 <- mutate(McAuliffe, age_log = log(Age))

ggplot(McAuliffe1, aes(x = age_log, y = PDI))+
  geom_point() +
  geom_smooth(method = "lm") +
  theme_light() +
  labs(title = "PDI Values by Soil Age; McAuliffe Profiles", x = "Soil Age (log scale)", y = "PDI Value")
```

```{r}
ggplot(McAuliffe1, aes(x = age_log, y = WM_PDI))+
  geom_point() +
  geom_smooth(method = "lm") +
  theme_light() +
  labs(title = "WM_PDI Values by Soil Age; McAuliffe Profiles", x = "Soil Age (log scale)", y = "WM_PDI Value")
```

Both graphs of the SRER data show a distinct peak around soil age of
650,000 years, about 6 on the log scale. Since PDI is a proxy for soil
development, we can infer that these soils are showing signs of
continuing development up until 650k years, before beginning to break
down again and returning to a more “undeveloped” state despite their
advanced age.

The McAuliffe graphs show minimal or no relationship between soil age
and PDI value. Several soils of the same age had highly different PDI
values.

Looking at the PDI index, signs of development include horizon
formation, color, texture, structure, consistence, presence of clay
films, and carbonate development. A soil that is very old has likely
been leached of most minerals, horizons are unclear, and erosion might
be occurring. However, erosion can occur for a variety of reasons, not
just soil age, as is likely evidenced in the McAuliffe profiles.

2)  Considering changes in PDI/WM PDI observed with erosion in the
    McAuliffe papers, do you think erosion plays any role in the
    observed patterns with age? Propose explanations for why SRER does
    or does not vary from the data from McAuliffe.

Yes, the PDI value for the Mid-Late Pleistocene site is 19.05, whereas
the eroded profile of the same age is 14.74. This is more apparent with
the Mid-Late Pleistocene lower site, with a PDI of 18.44; the eroded
profile is 2.85. Erosion seems to “turn back time” and revert a soil
back to an undeveloped state. I would guess that the SRER profiles were
not in areas that high erosion potential, and thus we could see a clear
relationship of soil age and PDI value.

***Component 3 - SRER Vegetation Transects***

These data were compiled for a series of long-term vegetation transects
monitored across the SRER. The transects are measured every 3 years and
span the entire elevation and soil age range across SRER. The data in
the spreadsheet represent the averages of all these data providing an
estimate of the annual aboveground native and invasive grass biomass
(g/m2), the total biomass (sum of native and invasive), the fraction of
the total biomass that is invasive, the average mean annual
precipitation (MAP, cm/yr), the surficial geologic mapunit (Qy, Qi3,
etc.), as well as ancillary data on pasture, burn history, grazing
history, UTM coordinates, and elevation (m). I subset the data to only
those sites that have not burned in recent history and are actively
grazed by cattle. \*\*Please note - I also included columns for MAP,
Total, Native and Invasive Biomass as Box-Cox transformed data (Name of
property followed by -X). This transformation helps approximate a normal
distribution for statistical analysis.

**1)Data coupling:** Update the vegetation dataset with the PDI and WM
PDI values from the spreadsheet with SRER and McAuliffe data. Assume all
the Qy mapunits are equivalent to "Qy1" from the chronosequence data.

```{r}
Total_SRER <- read.csv("../data/SRER_compiled.csv")

Total_SRER1 <- Total_SRER %>% 
  mutate(MapUnit_12 = replace(MapUnit_12, MapUnit_12 == "Qy", "Qy1"))

Total_SRER2 <- full_join(Total_SRER1, SRER1, join_by(MapUnit_12 == Site))

```

2)  **Data analyses:**

<!-- -->

i)  Create box plots of total, native, invasive biomass, and fraction
    invasive by Map Unit;

```{r}
rename_SRER <- Total_SRER2 %>% 
  rename(pasture = PASTURE, burn_status = BURN.STATUS, grazing_status = Grazing.Status, map_unit = MapUnit_12,
         utm_x = UTM_X, utn_y = UTM_Y, elevation_m = Elevation..m., map_mm_yr = Predicted.MAP..mm.yr.,
         avg_npg_biomass = Average.NPG.Biomass..g.m..., sd_npg_biomass = Std.Dev.NPG.Biomass,
         avg_ipg_biomass = Average.IPG.Biomass..g.m..., sd_ipg_biomass = Std.Dev.IPG.Biomass,
         totalgrass_biomass = Total.Grass.Biomass..g.m..., fraction_invasive_biomass = Fraction.Invasive.Biomass,
         boxcox_Map_mm_yr = Predicted.MAP..mm.yr..X, boxcox_avg_npg_biomass = Average.NPG.Biomass..g.m....X,
         boxcox_avg_ipg_biomass = Average.IPG.Biomass..g.m....X, boxcox_totalgrass_biomass = Total.Grass.Biomass..g.m....X) %>% 
  filter(!map_unit %in% c("Qi4", "Qy2"))

mapunit_order <- c("Qy1", "Qi3", "Qi2", "Qi1", "Qo")
rename_SRER$map_unit <- factor(rename_SRER$map_unit, levels = mapunit_order)

ggplot(rename_SRER, aes(x = map_unit, y = totalgrass_biomass)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.1) +
  labs(x = "Map Unit", y = "Total Biomass (g) of Invasive and Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = map_unit, y = avg_npg_biomass)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.1) +
  labs(x = "Map Unit", y = "Total Biomass (g) of Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = map_unit, y = avg_ipg_biomass)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.1) +
  labs(x = "Map Unit", y = "Total Biomass (g) of Invasive Grass")
```

```{r}
ggplot(rename_SRER, aes(x = map_unit, y = fraction_invasive_biomass)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.1) +
  labs(x = "Map Unit", y = "Fraction of Total Biomass (g) that is Invasive Grass")
```

Here we can see several trends that emerge. The x-axis is map unit and
is arranged by soil age from youngest soils to oldest soils. Total
biomass, invasive biomass, and fraction of invasive biomass are all
positive trends with soil age. Native biomass has a negative trend,
indicating that native grass cannot persist on older soils.

ii) Plot total, native and invasive biomass, and fraction invasive by
    mean annual precipitation;

```{r}
ggplot(rename_SRER, aes(x = map_mm_yr, y = totalgrass_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Average Annual Precipitation (mm)", y = "Total Biomass (g) of Invasive and Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = map_mm_yr, y = avg_npg_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Average Annual Precipitation (mm)", y = "Total Biomass (g) of Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = map_mm_yr, y = avg_ipg_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Average Annual Precipitation (mm)", y = "Total Biomass (g) of Invasive Grass")
```

```{r}
ggplot(rename_SRER, aes(x = map_mm_yr, y = fraction_invasive_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Average Annual Precipitation (mm)", y = "Fraction of Total Biomass (g) that is Invasive Grass")
```

Annual average precipitation has a positive relationship with all
variables. Native grass biomass does not increase nearly as much as
invasive grass biomass, indicating that native grasses do not respond as
aggressively to higher annual average precipitation.

iii) Plot total, native and invasive biomass, and fraction invasive by
     PDI and WM PDI.

```{r}
ggplot(rename_SRER, aes(x = PDI, y = totalgrass_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "PDI Value", y = "Total Biomass (g) of Invasive and Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = PDI, y = avg_npg_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "PDI Value", y = "Total Biomass (g) of Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = PDI, y = avg_ipg_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "PDI Value", y = "Total Biomass (g) of Invasive Grass")
```

```{r}
ggplot(rename_SRER, aes(x = PDI, y = fraction_invasive_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "PDI Value", y = "Fraction of Total Biomass (g) that is Invasive Grass")
```

```{r}
ggplot(rename_SRER, aes(x = WM_PDI, y = totalgrass_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "WM_PDI Value", y = "Total Biomass (g) of Invasive and Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = WM_PDI, y = avg_npg_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "WM_PDI Value", y = "Total Biomass (g) of Native Grass")
```

```{r}
ggplot(rename_SRER, aes(x = WM_PDI, y = avg_ipg_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "WM_PDI Value", y = "Total Biomass (g) of Invasive Grass")
```

```{r}
ggplot(rename_SRER, aes(x = WM_PDI, y = fraction_invasive_biomass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "WM_PDI Value", y = "Fraction of Total Biomass (g) that is Invasive Grass")
```

Much like the first set of graphs produced in this document of map unit
vs. biomass, we see similar relationships here. Total, invasive, and
fraction biomass increase with PDI and WM PDI values, although total
biomass is only slightly positive. Native biomass decreases slightly
with PDI and WM PDI values. Lehmann's lovegrass—the primary invasive
grass in this dataset—produces more biomass than native grasses, and is
also more prolific. This likely explains the positive trend in total and
fraction biomass.

3)  **Statistical analyses:**

<!-- -->

i)  Fit least square means linear regressions using MAP as the
    independent variable, and total, native and invasive biomass, and
    fraction invasive as dependent variables - \*\*\*use the Box-Cox
    transformed variables!

```{r}
library(emmeans)

totalgrass_model <- lm(boxcox_totalgrass_biomass ~ boxcox_Map_mm_yr, data = rename_SRER)

npg_model <- lm(boxcox_avg_npg_biomass ~ boxcox_Map_mm_yr, data = rename_SRER)

ipg_model <- lm(boxcox_avg_ipg_biomass ~ boxcox_Map_mm_yr, data = rename_SRER)

fx_model <- lm(fraction_invasive_biomass ~ boxcox_Map_mm_yr, data = rename_SRER)

summary(totalgrass_model)
summary(npg_model)
summary(ipg_model)
summary(fx_model)

library(sjPlot)
tab_model(totalgrass_model, npg_model, ipg_model, fx_model)

```

Here we can see that MAP is a highly significant driver of biomass for
all variables (p \< 0.01). The strength of this relationship varies
between native and invasive biomass. Total (Adjusted R\^2 = 0.331) and
invasive (adjusted R\^2 = 0.32) biomass showed strong responses to MAP
(p\<0.001). In contrast, native biomass is less responsive, as MAP
explains only 7.5% of variance (adjusted R\^2=0.075).

ii) Fit least square means linear regressions using estimated age, PDI,
    and WM-PDI as the independent variable, and total, native and
    invasive biomass, and fraction invasive as dependent variables -
    \*\*\*use the Box-Cox transformed biomass data!

```{r}
totalgrass_age_model <- lm(boxcox_totalgrass_biomass ~ age_log, data = rename_SRER)

npg_age_model <- lm(boxcox_avg_npg_biomass ~ age_log, data = rename_SRER)

ipg_age_model <- lm(boxcox_avg_ipg_biomass ~ age_log, data = rename_SRER)

fx_age_model <- lm(fraction_invasive_biomass ~ age_log, data = rename_SRER)

summary(totalgrass_age_model)
summary(npg_age_model)
summary(ipg_age_model)
summary(fx_age_model)

tab_model(totalgrass_age_model, npg_age_model, ipg_age_model, fx_age_model)
```

```{r}
totalgrass_PDI_model <- lm(boxcox_totalgrass_biomass ~ PDI, data = rename_SRER)

npg_PDI_model <- lm(boxcox_avg_npg_biomass ~ PDI, data = rename_SRER)

ipg_PDI_model <- lm(boxcox_avg_ipg_biomass ~ PDI, data = rename_SRER)

fx_PDI_model <- lm(fraction_invasive_biomass ~ PDI, data = rename_SRER)

summary(totalgrass_PDI_model)
summary(npg_PDI_model)
summary(ipg_PDI_model)
summary(fx_PDI_model)

```

```{r}
totalgrass_WMPDI_model <- lm(boxcox_totalgrass_biomass ~ WM_PDI, data = rename_SRER)

npg_WMPDI_model <- lm(boxcox_avg_npg_biomass ~ WM_PDI, data = rename_SRER)

ipg_WMPDI_model <- lm(boxcox_avg_ipg_biomass ~ WM_PDI, data = rename_SRER)

fx_WMPDI_model <- lm(fraction_invasive_biomass ~ WM_PDI, data = rename_SRER)

summary(totalgrass_WMPDI_model)
summary(npg_WMPDI_model)
summary(ipg_WMPDI_model)
summary(fx_WMPDI_model)
```

iii) Multiple linear regression - include MAP and the age variable with
     the best correlation (estimated age, PDI or WM-PDI) and their
     interaction term (MAP x Age) as the independent variables and
     total, native and invasive biomass, and fraction invasive as
     dependent variables - \*\*\*use the Box-Cox transformed variables!

| Relationship       | Adjusted R\^2 value |
|--------------------|---------------------|
| log_age x total    | .04                 |
| log_age x npg      | .06                 |
| log_age x ipg      | .14                 |
| log_age x fraction | .21                 |
| PDI x total        | .02                 |
| PDI x npg          | .07                 |
| PDI x ipg          | .01                 |
| PDI x fraction     | .18                 |
| WM x total         | .0007               |
| WM x npg           | .02                 |
| WM x ipg           | .05                 |
| WM x fraction      | .10                 |

The variable log_age has the most consistently high adjusted R\^2
values, so that is what I will use for the multiple linear regression
models.

```{r}
mlr_total <- lm(boxcox_totalgrass_biomass ~ map_mm_yr * age_log, 
                data = rename_SRER)

mlr_npg <- lm(boxcox_avg_npg_biomass ~ map_mm_yr * age_log, data = rename_SRER)

mlr_ipg <- lm(boxcox_avg_ipg_biomass ~ map_mm_yr * age_log, data = rename_SRER)

mlr_fx <- lm(fraction_invasive_biomass ~ map_mm_yr * age_log, data = rename_SRER)

summary(mlr_total)
summary(mlr_npg)
summary(mlr_ipg)
summary(mlr_fx)
```

**4) Summary and Conclusions**

Create figures and tables summarizing the data analysis and provide a
detailed discussion of the data patterns and propose explanations for
observed patterns. The discussion of the data should pull examples for
comparison from the published literature.

```{r}

#creates a table of model summaries
tab_model(mlr_total, mlr_npg, mlr_ipg, mlr_fx)

```

Results: Mean annual precipitation (MAP) is the dominant driver for
biomass (p \< 0.05) across all vegetation types. No significance was
found in interactions between soil age and MAP (p \> 0.05) in all
models. The model explained 45% of variance in invasive grass biomass
(adjusted R\^2 = 0.448) and only 16% in native grass (adjusted R\^2 =
0.161).

```{r}
long_SRER <- rename_SRER %>%
  pivot_longer(
    cols = c(boxcox_totalgrass_biomass, boxcox_avg_npg_biomass,
             boxcox_avg_ipg_biomass),
    names_to = "Biomass_Type", 
    values_to = "Value"
  ) %>% 
  mutate(Biomass_Type = recode(Biomass_Type,
    "boxcox_totalgrass_biomass" = "Total Grass Biomass",
    "boxcox_avg_npg_biomass"    = "Native Grass Biomass",
    "boxcox_avg_ipg_biomass"    = "Invasive Grass Biomass"
  ))

ggplot(long_SRER, aes(x = map_mm_yr, y = Value)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ Biomass_Type)
```

Discussion:

Here we have 2 datasets, a set from the Santa Rita Experimental Range
(SRER), and another set from the Tucson Mountains, Tortolitas, and the
Silverbell mountains. Both show general trends of PDI value increasing
with soil age, but with high variation in eroded sites (McAuliffe,
1994). While PDI values are generally seen as a proxy for soil age,
several soil profiles here are exceptions to this rule, so PDI alone
cannot be used as an accurate predictor of soil age. Another general
trend of soil age is vegetation type. As soils age, certain vegetation
can no longer persist there. We can see this in these soil profiles, as
native and invasive grass have an inverse relationship to PDI, with
invasive grass generally increasing with PDI value, and native grass
decreasing. This raises an interesting question, why can invasive grass
persist on older soil surfaces? Old soils are generally leached of
available minerals, and yet Lehmann's lovegrass is prevalent. Mean
annual precipitation (MAP) adds another piece to the puzzle, and is our
strongest predictor of grass biomass. Interestingly, native grass
respond less aggressively to higher MAP, perhaps indicating that they
are less dependent on moisture or more drought tolerant than invasive
grasses. The lack of significant interaction between MAP and soil age
suggests that these two variables act independently on biomass, and that
precipitation effect is stable across all vegetation types.

McAuliffe, J. (1994). Landscape Evolution, Soil Formation, and
Ecological Patterns and Processes in Sonoran Desert Bajadas. *Ecological
Monographs, 64*(2): 111-148.
